Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3): Appendices

### 2016-10-05 WSSSPE3/report/WSSSPE3_appendix.tex

\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)}

@@

\begin{abstract}
This report records and discusses the Third Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE3). The report includes a
description of the keynote presentation of the workshop, which served as an overview of sustainable
scientific software. It also summarizes a set of lightning
talks in which speakers highlighted to-the-point lessons and challenges
pertaining to sustaining scientific software. The final and main contribution of the report is a summary of the
discussions, future steps, and future organization for a set of self-organized
working groups on topics including developing pathways to funding scientific
software; constructing useful common metrics for crediting software
stakeholders; identifying principles for sustainable software engineering
design; reaching out to research software organizations around the world; and
building communities for software sustainability. For each group, we include a
point of contact and a landing page that can be used by those who want to join
that group's future activities. The main challenge left by the workshop is to
see if the groups will execute these activities that they have scheduled, and
how the WSSSPE community can encourage this to happen.

@@

\maketitle
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Third Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13;
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14; and
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational; and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged a broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future for software development and usage in science and
engineering. But WSSSPE2 did not have a good way to enact those mechanisms, or to
encourage the attendees to follow through on their intentions.

The WSSSPE3 workshop included multiple mechanisms for participation and
encouraged team building around solutions. WSSSPE3 strongly encouraged participation
of early-career scientists, postdoctoral researchers, graduate students,  
early-stage researchers, and those from underrepresented groups,
with funds provided to the conference organizers by the Moore Foundation, the
National Science Foundation (NSF), and the Software Sustainability Institute (SSI) to
support the travel of potential participants who would not otherwise be able to
attend the workshop. These
funds allowed 16 additional people to attend and participate.

WSSSPE3 also included two professional
event organizers/facilitators from Knowinnovation who helped the organizing committee members
plan the workshop agenda, and during the workshop, they actively engaged
participants with various
tools, activities, and reminders.

This report is based on collaborative notes taken during the workshop, which
were linked from the GitHub issues that represented the potential and actual
working
groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
Overall, the report discusses the organization work done before the workshop
(\S\ref{sec:preworkshop}), the keynote (\S\ref{sec:keynote}), and the lightning talks
presented at the meeting (\S\ref{sec:lightning}). The report also gives summaries of
action plans proposed by the working groups (\S\ref{sec:WGs}), 
then gives longer
descriptions of the activities that occurred in each of the working groups that
made substantial progress
(\S\ref{sec:group_details}), and provides some
conclusions (\S\ref{sec:conclusions}). The appendices contain lists of the organizing committee
(Appendix~\ref{sec:orgcom}), the registered attendees
(Appendix~\ref{sec:attendees}), and the travel award recipients
(Appendix~\ref{sec:awardees}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Calls for Participation} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting
a process to make progress in sustainable software, as the calls for
participation said:

\begin{quote} The WSSSPE1 workshop engaged the broad scientific community to
identify challenges and best practices in areas relevant to sustainable
scientific software. WSSSPE2 invited the community to propose and discuss
specific mechanisms to move towards an imagined future practice of software
development and usage in science and engineering. WSSSPE3 will organize
self-directed teams that will collaborate prior to and during the workshop to
create vision documents, proposals, papers, and action plans that will help the
scientific software community produce software that is more sustainable,
including developing sustainable career paths for community members. These teams
are intended to lead into working groups that will be active after the workshop,
if appropriate, working collaboratively to achieve their goals, and seeking
funding to do so if needed. \end{quote}

The first call for participation requested lightning talks, where each author
could make a brief statement about work that either had been done or was needed,
with the goal of contributing to the discussion of one or more working groups.
There were~24 lightning talks submitted; after a peer-review process, 16 
were accepted, as discussed further in Section~\ref{sec:lightning}.

The first call also discussed the potential action topics that came out of
WSSSPE2, and requested additional suggestions. The combination of existing and
new topics led to the following 18 potential topics that were advertised in the
subsequent calls for participation:
%
%\begin{quote}
\begin{itemize}
\renewcommand{\labelenumi}{\textbf{\theenumi}.}
\setlength{\rightmargin}{1em}

\item Development and Community
\begin{itemize}
\item Writing a white paper/review paper about best practices in developing
sustainable software
\item Documenting successful models for funding specialist expertise in software
collaborations
\item Creating and curating catalogs for software tools that aid sustainability
(perhaps categorized by domain, programming languages, architectures, and/or
functions, e.g., for code testing, documentation)
\item Documenting case studies for academia/industry interaction
\item Determining effective strategies for refactoring/improving legacy
scientific software
\item Determining principles for engineering design for sustainable software
\item Create a set of guidance giving examples of specific metrics for the
success of scientific software in use, why they are chosen, what they are
useful to measure, and any challenges/pitfalls; then publish this as a white
paper
\end{itemize}

\item Training
\begin{itemize}
\item Writing a white paper on training for developing sustainable software, and
coordinating multiple ongoing training-oriented projects
\item Developing curriculum for software sustainability, and ideas about where
such curriculum would be presented, such as a summer training institute
\end{itemize}

\item Credit
\begin{itemize}
\item Hacking the credit and citation ecosystem (making it work, or work better,
for software)
\item Developing a taxonomy of contributorship/guidelines for including software
contributions in tenure review
\item Documenting case studies of receiving credit for software contributions
\item Developing a system of awards and recognitions to encourage sustainable
software
\end{itemize}

\item Publishing
\begin{itemize}
\item Developing a categorization of journals that publish software papers
(building on existing work), and case studies of alternative publishing
mechanisms that have been shown to improve software discoverability/reuse, e.g.,
popular blogs/websites
\item Determining what journals that publish software paper should provide to
their reviewers (e.g., guidelines, mechanisms, metadata standards)
\end{itemize}

\item Reproducibility and Testing
\begin{itemize}
\item Building a toolkit that could allow conference organizers to easily add a
reproducibility track
\item Documenting best practices for code testing and code review
\end{itemize}

\item Documentation
\begin{itemize}
\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the
community to easily find up-to-date information on a WSSSPE topic (e.g.,
software credit, scientific software metrics, testing scientific software).
\end{itemize}

\end{itemize}
%\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote}\label{sec:keynote}
%\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
Department of Astronomy, University of Illinois, titled \emph{Why Sustain
Scientific Software?}. Turk is a prolific scientific software practitioner and
has extensive experiences working on large collaborative projects employing
modern computing tools \cite{2011ApJS..192....9T}. He also co-organizes and
champions WSSSPE events.

In his keynote address, Turk recapped the course of development of WSSSPE
workshops over the past few years, alongside his career development from a
postdoc to an academic. The first WSSSPE workshop was at the Supercomputing
conference (SC13) in 2013, but he observed that the notion of sustainable
scientific software drew in an audience beyond supercomputing. In the following
year, WSSSPE1.1 at SciPy had speakers talking about how software has been
sustained inside the scientific Python community. WSSSPE2 at SC14 had breakout
group discussions coming up with actionable items, and WSSSPE2.1 at SciPy 2015
was similar. Turk noted the different atmosphere of the surrounding large
conferences, despite similar WSSSPE participants.

WSSSPE3 left the traditional Supercomputing Conference environment this year,
and in Turk's words, this change spoke to the fact that scientific software
comes from many different types of inquiries, deployment, strategies for
maintenance, users, and ways of measuring the value of a piece of software. It
appeared to Turk that the supercomputing community generally adopts some
top-down approaches, whereas the SciPy community more often than not uses more
bottom-up systems. According to Turk, there is a divergence in views about
progress in software: the
supercomputing community thinks that software is getting harder, with exascale
computing and optimization issues in mind; but the SciPy community thinks that
software is becoming better, with emerging tools such as Jupyter and
productivity packages for research workflows. Admitting such comparisons are
somewhat unfair generalizations, Turk reminded the audience that the different
approaches bring different types of ideas to the table, and he welcomed WSSSPE3
being conducted outside existing preconceptions.

Returning to the topic of his talk, Turk invited the audience to picture
scientific software as a flower on a landscape under the Sun, which may
represent a number of measurable factors such as number of citations; growth of
a community and number of contributors; amount of funding; prestigious prizes
awarded; stability of the community in terms of leadership transitions, serving
community needs, not breaking test suites, and performance on new architectures.
But all these metrics are strictly speaking \emph{proxies} for the values and
the impact scientific software bears. What we can measure does not give us
direct insight---it just gives us proxies of insight.

Turk then moved onto various different definitions of sustainability. His
favorite one was ``keeping up with bug reports,'' where even if no new features
were added, the software remains sustainable. Another definition of
sustainability Turk mentioned was ``adding of new features,'' or ``maintaining
the software for a long period of time'' such as the cases of \TeX\ or \LaTeX\
with community help. A notion Turk heard often at supercomputing conferences was
that sustainable software ``continues to work on new architectures.'' Yet
another metric was ``people continuing to be able to learn how to use and apply
the software.'' A funder Turk heard talked about sustainability as ``continuing
to get funded.'' Turk also recalled that Greg Wilson, among others, said in
WSSSPE1.1 that his view of sustainable software was software that ``continued to
give the same results over time.'' A last measure of sustainability Turk
presented was ``the ability to transition between different people developing
and using a piece of software.''

At WSSSPE1, several models were presented for ensuring sustainability. Turk
considered that a familiar one was a funded piece of software where an external
agency provided funds to a group who are not necessarily exclusively working on
and developing the software, keeps it going, and provides it to the scientific
community. The model of productized software, in which a piece of software has
grown to the point that research groups or people are willing to support it with
some amount of funding, for instance, a subscription to use cloud services that
deploy a piece of software, or purchase of a piece of software. A final model
Turk felt conflicted about is a volunteer model that is traditional
old-school---not modern-day open source---development.

Turk discussed whether productizing scientific software was synonymous with
being sustainable and self-sufficient. He thought it was not necessarily the
case and furthermore, it could lead to a divergence of interests between users
and developers.

Turk reminded the audience that the volunteer model means unpaid labor. On this
note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
the open source software
community\footnote{\url{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}}.
Oftentimes, a person funded to work full time on a scientific project can spend
a small amount of time for working on a piece of software necessary for that
project. However, researchers' abilities to participate in that volunteer
community are not always the same and may not always be aligned with their
research projects. From Turk's experience, we cannot always rely on unpaid labor
and volunteer time to sustain a piece of software---this came down to the
notions of the top-down and the bottom-up approaches, i.e., the funded versus
the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
projects can be just as large-scale as a top-down software development project.

Turk said that sustaining scientific software really meant to him conducting
scientific inquiries, often by some specific software, and sustaining the people
we care about, our careers, and the future of our fields. According to Turk, we
all have an invested stake in sustaining scientific software. Hence, having
``sustained'' projects can suffocate new projects, so we need to make sure we
don't cause novel ideas and packages to suffer at the hands of the status quo.

Turk talked about possible reasons why we want to sustain scientific software:
devotion to science and interests in pursuing the next stage of research; fun
and creative thrill in writing codes and papers; usefulness with measurable
impacts, for example, LINPACK and HDF groups providing data storage to
satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
his wishlist of questions to be answered in the future:
%
\begin{itemize}

\item How do we ship a product on time when dealing with a mix of funding models
and motivations especially when we rely on volunteers?

\item How do we know when it is time to end some software and move on? For
example, should we stop sustaining Python and switch to Julia and Javascript?

\item How can productized software balance its future versus its past, or the
emerging needs of the customers versus the existing needs of the development
community?

\item How can we help avoid burnout and retain the joy in the communities?

\item How can we reduce systemic bias, which goes back to Dryden's blog post
especially on how ethics of unpaid labor disproportionately affect
underrepresented communities?

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning Talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}{Slides.}}
\end{comment}

After the keynote, WSSSPE3 continued with lightning talks.
These short talks were intended to give an opportunity for attendees to
quickly highlight an important issue or a potential solution.
%
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
Facilitating Research Training While Supporting Scientific Research}}. Benjamin
Tovar presented a case study of the Cooperative Computing Lab (CCL)\footnote{\url{http://ccl.cse.nd.edu}} at the
University of Notre Dame, which is a small group of individuals whose main tasks
are collaborating with people that have large-scale computing problems,
operating various parallel computer systems, conducting computer science
research, and developing open source software. One of the main challenges they
face is finding a balance between flexibility/training and stability/quality.
Their current solution for ensuring the latter was to add a software engineer
(the presenter) to the existing team of faculty and students, who now also
serves as a ``spring'' between flexibility and stability.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}}. The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it is not well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly.
Instead, sustainability at its heart is a systemic concept and must be viewed
from a range of different dimensions including environmental, economic,
individual, social, and technical. She introduced the Karlskrona Manifesto on
Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
of related work on the topic of sustainability into a set of (mis-)perceptions
and principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
to Improve Productivity of Legacy Tools}}. Abani Patra presented an example of
how the community using Titan2D, a geoflow simulation software, increased the
productivity of their tools by improving both code and data layout~\cite{Patra20051}. The main
obstacles in this change were the non-existence of a common version control
system for the source code, coupled with multiple versions of the same code
base, the fixed format of input files, that many input values were set as
compilation flags, and that the internal data layout was not suitable for modern
technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
developers included reinforcing the code structure using multiple layers of
Python and C++ interfaces, and a redesign of the data layout to be more suitable
for modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software:
Lessons from Open Source}}. Abigail Cabunoc Mayes combines two properties of open
source that, together, create a suitable habitat for sustainable software. The
first of the two properties, public, does not only mean public code. It also
includes public discussions, a public process of including contributions, and an
open license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability, and ease of understanding, but also
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}}. Lorraine Hwang presented experiences with the Computational
Infrastructure for Geodynamics (CIG)\footnote{\url{http://geodynamics.org/}}, a community software with a worldwide user
base. Like others, their main goals include high usability, sustainability, and
reproducibility. As a means to achieve these goals, various communication
channels have been developed, such as mailing lists, wikis, workshops,
hackathons, tutorials, and webinars. In order to contribute to the
infrastructure, codes must adhere to specified minimum standards with the desire
that all codes are working toward target standards.
%, but ideals are also given to encourage reaching the aforementioned goals. 
These include, e.g., the use of version control, certain coding styles, the
presence and nature of code tests and documentation, and certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}}. In a second talk, Lorraine
Hwang mentioned a variety of ways to cite efforts within the SAGA framework\footnote{\url{https://geodynamics.org/cig/projects/saga/}}, including
science papers, code papers, user manuals, and the CIG website. An analysis of
the resulting citations showed that $80\%$ of papers that use CIG codes mention
the code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better cited. In
part, this is attributed to the fact that CIG requires that donated software
provide a citable paper specified in the User Manual. The project is working on
tools and methods to generate attribution information automatically.
%This,
%in part, is attributed to suggested citation forms communicated to the
%community, and runtime tools that help collecting a relevant citation list
%automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}}. DASPOS\footnote{\url{http://daspos.org}} is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to
other disciplines such as biology, astrophysics, and digital curation. It includes
physicists, digital librarians, as well as computer scientists, and aims to
achieve some commonality across disciplines. Examples are meta-data descriptions
of archived data, computational descriptions, descriptions of how data was
processed, questions such as whether computation replication can be automated,
and what the impact of access policies on the preservation infrastructure is.
One of the products of this effort is a suite of tools that deals with this
preservation~\cite{Meng2015137,Meng:2015:UPE:2755979.2755982,1742-6596-664-3-032022,nd4},
and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user community
from the start, and to budget that way. For the specific example of the
preservation software, this means that besides adding value to the community, it
needs to be transparent to their workflows, i.e., not requiring additional
effort to preserve ``research objects.''

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}}. James
Hetherington started by listing some of the problems research software faces,
which include poor standard of verification and low levels of reuse. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar~\cite{rse1,rse2}. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who do not want to serve as programmers, and
creating synergies with research computing platforms. The success of such a
group would be measured by the output and quality of the research software.
Members could be part of Research Computing or faculty. They would not be
independent researchers, but would have to have a research background. An
attempt to form a community of RSE groups within the UK has been underway for
several years, including funding from the UK Research Council. An open question
is whether this approach can be adopted in other countries, including the~USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}}. Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, proficient developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late interaction
with users. Instead, an alternative approach was proposed that, at first, skips
the development phase and repeatedly goes through requirement gathering, design
and user interaction/learning phases, and only eventually starts development
once an agreement is reached, leading to an increased user satisfaction, higher
adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
Platform for Predictive Biology: Lessons Learned on the KBase Project}}. 
Dan Gunter presented experiences and lessons learned as part of KBase\footnote{\url{http://kbase.us}}, an
open software and data platform for addressing the grand challenge of systems
biology: predicting and designing biological function. KBase is a unified system
that integrates data and analytical tools for comparative functional genomics of
microbes, plants, and their communities. However, it is also a collaborative
environment for sharing methods and results, and placing those results in the
context of knowledge in the field. Being a large, multi-institutional project,
one of the big challenges is to agree on standards to enable a single,
maintainable system. Working in isolation does not work (anymore) within this
field, and the community in the field also does not have standards for software
engineering. This is contrasted to computer science research, where software
engineering standards shorten design cycles, leading to more time for highly
rewarded activities like publishing, performance studies, graduating students,
or protecting ideas before publication. Instead of this more traditional
approach, KBase uses a variation on the ``Scrum'' methodology. After picking
projects and team members, four to five teams work on projects for about two
weeks before a one-week evaluation by the executive committee. Based on their
assessment new teams and projects might be chosen before iteratively restarting
the ``agile'' development cycle. This process is intentionally open and
documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}}. Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects among
experimental data, research software, and publications. A current effort in the
Geosciences is the ``Geoscience Paper of the Future,''\footnote{\url{http://www.ontosoft.org/gpf}} which includes four
elements~\cite{ESS2:ESS284}. First, it forms a modern paper including text, data, and pointers to
supplementary materials. Second, it is reproducible, including data processing,
workflow, and visualization tools. Third, it is part of open science, which
includes being publicly available under open licenses, and providing meta-data.
Last but not least, it uses digital scholarship elements like persistent
identifiers for software, data, and authors, and it cites both data and
software. OntoSoft is a tool for helping with this effort, providing software stewardship
for the Geosciences. As part of this initiative, a special issue of a journal in
Geoscience areas is planned to include only Geoscience papers of the future,
with submissions open until the end of 2015~\cite{GPFI}. In addition, training sessions are
provided to geoscientists to learn best practices in software and data sharing,
provenance documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
Accreditation Framework}}. Neil Chue Hong presented a proposal to build a
scientific software accreditation framework. One of the aims of such a framework
would be to measure how ``good'' a given piece of software is, and to evaluate
how this can be effectively measured in the first place. This can be compared to
the effort of the standardized, easy to read, and understandable nutritional
labeling of food, which only contain a small set of categories. However, such a
framework for software would be more difficult due to different existing
community norms. The challenges such a framework faces include that many
measurements are subjective, that many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
Support for Sustainable Scientific Software}}. Carver argued that for scientific
software to be truly sustainable, there is a need for developers to use
appropriate software engineering practices. His experience interacting with
scientific teams indicates that choosing and tailoring these practices is not a
trivial exercise~\cite{Nanthaamornphong-Carver:2015, Heaton-Carver:2015, Basili-etal:2008, 4222616}.
There is a general culture clash between software engineering
and science that hinders our ability to communicate and choose appropriate
methods. In addition, many experienced scientific software developers appear to
be unaware of software engineering practices that may be beneficial to them. The
most appropriate software engineering practices are those that are lightweight,
properly tailored, and focus on the key software development problems faced by
scientists. In order to increase the use of software engineering in science, we
need more documented success stories. These successes need to be socialized 
%\choinote{not sure if ``socialize'' can be used in such a way}
within the scientific community through workshops like the Software Engineering
for Science workshop
series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
and the new Software Engineering track in \textit{Computing in Science and
Engineering} magazine.

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}.
This talk highlighted an attempt to solve the common problem for open source
development: it is difficult to collect information about how many people use
particular software, how often, which version, which parts of the software, or
on which operating system. Current solutions include surveys, but these have
high uncertainties. A different approach is based on automatic ``call-backs''
that collect these information at runtime and send it to a central place for
analysis. Problems with this approach include obtaining agreement from the user,
legal issues, increased maintenance (of servers), ethical questions, and also the
lack of a common infrastructure. Some of these problems are of a social nature
and have to be solved as such, but the last problem (a missing common
infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
which is developing common APIs and a library implementation for common,
repeating tasks such as obtaining user consent. Results are uploaded not
directly to project servers, but to neutral third parties that only publish
aggregated statistics to projects.

\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
included)}}. Alice Allen described the Astrophysics Source Code
Library\footnote{\url{http://ascl.net}}, an increasingly used way to obtain a unique ID for astrophysics
software that is indexed by indexing services and can be cited~\cite{wssspe2_allen_jors}.
ASCL offers
clones of existing infrastructure, provides server space and computing
resources, shares innovations, and permits moves elsewhere. Users provide a
domain name, then control and configure their site and use it as intended,
gather their codes as they wish, share innovations, and protect the provided
computing environment.

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
Developing Community}}. The effort behind this talk given by Ray Idaszak started
from a breakout group at an NSF SI2 workshop in 2015, and centers around
building a framework for creating metrics for scientific software~\cite{ahalt-csessp}. This
framework would improve both the metrics and the software it evaluates, and
could also serve as a tool for building a community around the idea. With
especially the last-mentioned idea (building a community) in mind, a software
``peer review group'' would be created, representing stakeholders who will
self-review software created by their respective communities, and will
concurrently develop metrics. The whole project should be community-governed,
without a single institution overseeing the activities or infrastructure, with
the hope to evolving community-generated and adopted standards. The generation
of metrics would be tied to the actual evaluation of software, creating an
incentive by improving the evaluated software itself during this process. The
framework code would provide infrastructure for the creation of metrics and
evaluation, and forums for generation of software success metrics. It would also
support code reviews of the evaluated software. An open question is whether it
is possible to fit the resulting metrics in a common template. So far, this is
still in a design phase, with a white paper at the CSESSP workshop 2015 and this
talk, but the WSSSPE workshops are seen as a forum for the community to assemble
and act, and is planned to be used also in the future to build this community
and framework.

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Working Groups: Summaries} \label{sec:WGs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

After the keynote and lightning talks, the workshop facilitators led an exercise to create working
groups for the rest of the meeting.  The
attendees first suggested additional topics beyond those in the call for participation (as listed in
\S\ref{sec:preworkshop}).  One topic that was introduced at this point was ``Building Sustainable
User Communities for Scientific Software.''  The full set of topics was then placed on
flipchart-sized pieces of paper around the walls, and the attendees voted on which topics
they were strongly interested in working on at WSSSPE3, and which topics they were 
generally interested in contributing to, but not as strongly.  This led to some topics being
taken out of the mix for the rest of the workshop, since not enough people wanted to contribute
to them to lead to a useful discussion.  Additionally, some topics were combined by the
participants who felt they were closely linked and that a group could address multiple of them
in a single discussion.

After this, the attendees broke up into small working groups to discuss
the remaining topics during most of the remaining 1 1/2 days.  A high-level summary
of each topic and group's work can be found in the subsections of this section, and for
all but one group, 
more detailed notes on each group's discussions can be found in \S\ref{sec:group_details}.

Midway through the first afternoon and between the two days, each group
had a chance to talk about what progress they had made.  As discussed below (in \S\ref{sec:legacy}), the group that formed to discuss Legacy Software dissolved after the first session
as group members left to join other groups.
In addition, in the morning of the
second day, a small set of reviewers/advisors (external to the groups but chosen by the group
members) visited each group to listen to what they group was planning and to provide feedback.

%\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}

\input{wg_main_best_practices}
\input{wg_main_funding_spec_expert}
%\input{wg_main_catalogs}
\input{wg_main_industry_interaction}
\input{wg_main_legacy_SW}
\input{wg_main_eng_design}
\input{wg_main_metrics}
\input{wg_main_training}
\input{wg_main_software_credit}
\input{wg_main_publishing_SW}
\input{wg_main_user_community}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Working Groups: Details} \label{sec:group_details}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section captures detailed reports from each working group that made significant
progress.  Each subsection records the discussion of
a group, as written by that group at that time (in the first person and in the present\slash future tense.)
Thus, the subsections are records of what the groups did and planned, as of the end of the WSSSPE3 workshop.

\input{wg_appendix_best_practices}
\input{wg_appendix_funding_spec_expert}
%\input{wg_appendix_catalogs}
\input{wg_appendix_industry_interaction}
%\input{wg_appendix_legacy_SW}
\input{wg_appendix_eng_design}
\input{wg_appendix_metrics}
\input{wg_appendix_training}
\input{wg_appendix_software_credit}
\input{wg_appendix_publishing_SW}
\input{wg_appendix_user_community}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3): Appendices}

@@

\begin{abstract}
This document contains three appendices to the report on the Third Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE3): Appendix A, The list of organizers; Appendix B, the list of
attendees; and Appendix C, the list of travel award recipients.

@@

\maketitle
